---
title: "STA 471 - Homework 9"
author: "Richard McCormick"
date: '2023-12-03'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# 1. Problem \textbf{T} on page 325 of the textbook.

```{r}
X <- c( 18, 20, 24, 28, 30, 33, 36, 48, 60, 40, 42, 45, 62, 71, 75 )
Y <- c( 4.8, 5.5, 5.8, 6.0, 6.5, 6.6, 6.7, 7.0, 7.3, 3.3, 3.8, 4.1, 5.0, 5.5, 6.0 )
D <- c( 0, 0, 0, 0, 0, 0, 0, 0, 0, 1, 1, 1, 1, 1, 1 )

hw9.data <- data.frame( Y, X, D )
```

## a. Define a suitable numerical dummy variable *D* to seperate the two groups(*f* and *m*) of data. Fit the model $Y = \beta_0 + \beta_1X + \beta_2D + \beta_3DX + \epsilon$ and give (1) the fitted equation, (2) the fitted line for each group (*f* and *m*) based on the fitted equation.

```{r}
hw9.model <- lm( Y ~ X + D + I( D * X ) )
summary( hw9.model )
```

### 1. Give the Fitted Equation   
The fitted equation for the model is:  
$Y = 4.518222 + 0.052310X - 3.640652D + 0.014659DX$

### 2. Give the Fitted Line for each group (*f* and *m*) based on the fitted equation.  
The fitted line for \textbf{f} is:  
$Y = 4.518222 + 0.052310X$  

The fitted line for \textbf{m} is:  
$Y = 0.877572 + 0.066969X$

## b. Test to determine whether the two straight lines are parallel at $\alpha = 0.05$.

### I. Hypothesis
$H_0: \beta_3 = 0$. The lines are parallel.  
$H_A: \beta_3 \neq 0$. The lines are not parallel.

### II. Test Statistics
Test Statistic: $t = \frac{b_3 - \beta_{3}}{se(b{3})}$  
Observed Statistic: $t = 1.188$

Critical T-Value ($t_{15-3-1} = t_{11}$): 2.201

### III. Conclusion
As $t_{obs} = 1.188 < 2.201 = t_{11}$, we can \textbf{accept} the null hypothesis and conclude that the lines are parallel with each other.

## c. Test to determine whether the two straight lines are coincident at $\alpha = 0.05$.

### I. Hypothesis
$H_0: \beta_2 = \beta_3 = 0$. The lines are coincident.  
$H_A: At\ least\ one\ of \beta_2, \beta_3 \neq 0$. The lines are not coincident.

### II. Test Statistics
Test Statistic: $F = \frac{MS_{reg}}{MS_{resid}}$
Observed Statistic: $F_{obs}$ = 62.85  

Critical f-Value ($f_{3, 11}$): 2.66

### III. Conclusion
As $f_{obs} = 62.85 > 2.66 = f_{3,11}$, we can \textbf{reject} the null hypothesis and accept the alternative hypothesis that the lines are not coincident.  

## d. Fit seperate straight-line regressions of *Y* on *X* for *f* and *m*, respectively. How do the fitted lines here compare to the fitted lines in part \textbf{a}?

```{r}
hw9.data.f <- subset( hw9.data, D == 0 )
hw9.model.f <- lm( Y ~ X, data=hw9.data.f )
summary( hw9.model.f )
```
The fitted model for *just* \textbf{f} is:
$Y = 4.518222 + 0.052310X$  
The model in part \textbf{a} was:
$Y = 4.518222 + 0.052310X$  

The two models are identical.

```{r}
hw9.data.m <- subset( hw9.data, D == 1 )
hw9.model.m <- lm( Y ~ X, data=hw9.data.m )
summary( hw9.model.m )
```

The fitted model for *just* \textbf{m} is:
$Y = 0.877570 + 0.066969X$  
The model in part \textbf{a} was:
$Y = 0.877572 + 0.066969X$  

The two models are identical.

