---
title: "STA471 - Homework 6"
author: "Richard McCormick"
date: '2023-11-09'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library( ggplot2 )
library( alr4 )
library( EnvStats )
library( dplyr )
```

## 1. Using least squares procedures, estimate the b’s in the model:
$Y$ = $\beta_0X_0 + \beta_1X_1+\beta_2X_2+\epsilon$

```{r}
X1 <- c( 1,4,9,11,3,8,5,10,2,7,6 )
X2 <- c( 8,2,-8,-10,6,-6,0,-12,4,-2,-4 )
Y <- c( 6,8,1,0,5,3,2,-4,10,-3,5 )

data <- data.frame( Y, X1, X2 )
model <- lm( Y ~ X1 + X2 )
summary( model )

print( paste( "b1 = ", coef( model )[2] ) )
print( paste( "b2 = ", round( coef( model )[3], 3 ) ) )
```

\newpage
## 2. Write out the analysis of variance table.

```{r}
anova( model )

RSS <- sum( anova( model )[1:2, 2] )
print( paste( "Regression Sum of Squares:", RSS ) )

TSS <- sum( anova( model )[,2] )
print( paste( "Total Sum of Squares:", TSS ) )

```

\newpage
## 3. Using $\alpha$ = 0.05, test to determine if the overall regression is statistically significant.

\textbf{I. Hypothesis}\newline
$H_0$: $\beta_1$ = $\beta_2$ = 0 \newline
$H_A$: At least one of: $\beta_1$, $\beta_2$ $\neq$ 0 \newline

```{r}
summary( model )
```

\textbf{II. Test Statistic}  
Test Statistic: $F$ = $\frac{MS_{reg}}{MS_{resid}}$ \newline
Observed Statistic: $F_{obs}$ = 7.176, from Summary Table. \newline
p-value = 0.01641, from Summary Table. \newline

\textbf{III. Conclusion}  
P-value = 0.01641 < $\alpha$ = 0.05.  
At the $\alpha = 0.05$ level of significance, the overall regression model is statistically significant. Thus, we \textbf{reject} the null hypothesis, and accept the alternative hypothesis.

\newpage

## 4. Calculate the square of the multiple correlation coefficient, namely, R2. What portion of the total variation about images is explained by the two variables?

```{r}
summary( model )

print( paste( "R-squared value is:", summary( model )$r.squared ) )
```
The portion of total variation about images explained by the two variables in this model is 64.21%

\newpage
## 5. The inverse of the X'X matrix for this problem is as follows:

$\begin{bmatrix}
4.3705 & -0.8495 & -0.4086\\
-0.8495 & 0.1690 & 0.0822\\
-0.4086 & 0.0822 & 0.0422\\
\end{bmatrix}$

```{r}
xinv_mat <- matrix( c( 4.3705, -0.8495, -0.4086, -0.8495, 0.1690, 0.0822,
                       -0.4086, 0.0822, 0.0422 ), 3, 3 )
```

## Using the results of the analysis of variance table with this matrix, calculate estimates of the following:

## a. Variance and confidence intervals of b1.

```{r}
# 1. Find s^2 = RSS / (n - 2)
s.squared = ( deviance(model) / 8 )

# 2. Create variance matrix, using (X'X)^-1
var_mat <- s.squared * xinv_mat

# 3. Get the i-1th element
b1.variance <- var_mat[2, 2]

print( paste( "Variance of b1 =", b1.variance ) )


# 95% confidence interval for b1
confint( model, level=0.95 )[2,]
```

## b. Variance and confidence intervals of b2.

```{r}
# Get the i-1th element
b2.variance <- var_mat[3,3]

print( paste( "Variance of b2 =", b2.variance ) )

# 95% confidence interval for b2
confint( model, level=0.95 )[3,]
```

\newpage
## 6. How useful is the regression using $X_1$ alone? What does $X_2$ contribute, given that $X_1$ is already in the regression?

```{r}
writeLines( paste( round( ( anova( model )[1,2] / RSS )*100, 3 ), 
            "% of the variation in Y is explained by the regression using X1 alone." ) )

model.X1 <- lm( Y ~ X1 + X2 )
anova( model.X1 )
RSS.X1 <- sum( anova( model.X1 )[1, 2] )
SS.X1 <- RSS - RSS.X1

writeLines( paste( round( ( SS.X1/RSS )*100, 3 ), 
              "% of the variation in Y is explained by the regression using X2,\n", 
              "given that X1 is already in the model." ) )
```

\newpage
## 7. How useful is the regression using $X_2$ alone? What does $X_1$ contribute, given that $X_2$ is already in the regression?

```{r}
writeLines( paste( round( ( anova( lm( Y ~ X2 ) )[1,2] / RSS ) * 100, 3 ), 
              "% of the variation in Y is explained by the regression using X2 alone." ) )

model.X2 <- lm( Y ~ X2 + X1 )
anova( model.X2 )
RSS.X2 <- sum( anova( model.X2 )[1, 2] )
SS.X2 <- RSS - RSS.X2

writeLines( paste( round( ( SS.X2 / RSS ) * 100, 3 ), 
              "% of the variation in Y is explained by the regression using X1,\n", 
              "given that X2 is already in the model." ) )
```

\newpage
## 8. What are your conclusions?

Given the model and our variables, it is reasonable to conclude that $X_1$ alone contributes the most to the explanation of variation in Y in this model, with more than 95% of total variation being explained. The p-value for $X_2$ is very high, which does not support a conclusion that $X_2$ contributes much to the model. The model would be more accurate when only using $X_1$, as $X_2$ does not contribute much to the overall regression.

### a. Fit an appropriate model to the data using $\alpha$ = 0.05 and compare the effectiveness of the appropriate model with the full model by adjusted $R^2$.

```{r}
fit.model <- lm( Y ~ X1 )
summary( fit.model )
summary( model )
```
The adjusted R-squared value is higher when only using $X_1$ - 0.5677 for the more appropriate model, and 0.5526 for the full model.